Flow-Based Cost Allocation in Linear Optimized Power Systems are non-unique but localizable

### 2020-08-13 paper/flow_based.tex
further work on chapter storage, seems good now

% general symbols
\newcommand{\state}{s_{i,t}}
\newcommand{\costfactor}{c_{i,t}}

@@

% storage
\newcommand{\storage}{g_{r,t}}
\newcommand{\storagedispatch}{\storage^\text{dis}}
\newcommand{\storagecharge}{\storage^\text{sto}}
\newcommand{\storagesoc}{\storage^\text{ene}}

@@

\newcommand{\operationalpricestorage}{o_r}
\newcommand{\capitalpricestorage}{c_r}
\newcommand{\capacitystorage}{G_r}
\newcommand{\capacitystorageupper}{\bar{G}_r}
\newcommand{\capacitystoragelower}{\ubar{G}_r}
\newcommand{\mulowerstorage}{\ubar{\mu}_{r,t}}
\newcommand{\muupperstorage}{\bar{\mu}_{r,t}}
\newcommand{\mulowerstoragedispatch}{\ubar{\mu}_{r,t}^\text{dis}}
\newcommand{\muupperstoragedispatch}{\bar{\mu}_{r,t}^\text{dis}}
\newcommand{\mulowerstoragecharge}{\ubar{\mu}_{r,t}^\text{sto}}
\newcommand{\muupperstoragecharge}{\bar{\mu}_{r,t}^\text{sto}}
\newcommand{\mulowerstoragesoc}{\ubar{\mu}_{r,t}^\text{ene}}
\newcommand{\muupperstoragesoc}{\bar{\mu}_{r,t}^\text{ene}}
\newcommand{\muupperstoragenom}{\bar{\mu}^\text{nom}_r}
\newcommand{\mulowerstoragenom}{\ubar{\mu}^\text{nom}_r}
\newcommand{\mustateofcharge}{\lambda^\text{ene}_{r,t}}

@@

% allocation quantities
\newcommand{\allocategeneration}[1][s, n]{A_{#1,t}}
\newcommand{\allocatestoragedispatch}[1][r, n]{A_{#1,t}}
\newcommand{\allocatepeer}[1][m \rightarrow n]{A_{#1,t}}
\newcommand{\allocateflow}[1][n]{A_{\ell,#1,t}}
\newcommand{\allocatetransaction}[1][s \rightarrow n]{A_{#1,\ell,t}}
\newcommand{\allocatestate}[1][i, n]{A_{#1,t}}
\newcommand{\allocatecost}[1][n \rightarrow i]{\mathcal{C}_{#1, t}}
\newcommand{\allocatecapexgeneration}[1][n \rightarrow s]{\mathcal{C}^{G}_{#1,t}}
\newcommand{\allocatecapexflow}[1][n \rightarrow \ell]{\mathcal{C}^{F}_{#1,t}}
\newcommand{\allocateopex}[1][n \rightarrow s]{\mathcal{O}_{#1,t}}
\newcommand{\allocateemissioncost}[1][n \rightarrow s]{\mathcal{E}_{#1,t}}

@@

In long-term operation and investment planning models, the total costs $\totalcost$ of a power system is the sum of mutliple cost terms. Typically, these include operational expenditures (OPEX) $\opexgeneration$ for generators, expenditures for emissions $\emissioncost$, capital expenditures (CAPEX) for generators $\capexgeneration$, CAPEX for the transmission system $\capexflow$ and possible other terms, \ie
\begin{align}
\totalcost &= \opexgeneration + \emissioncost +  \capexgeneration +  \capexflow + ...
\label{eq:total_cost}
\end{align}
In turn, each single cost term on the right side $\mathcal{C} = \{\opexgeneration, \emissioncost, \capexgeneration, \capexflow, ...\}$ consists of cost associated to the asset $i$ in the system, 
\begin{align}
    \cost = \sum_{i} \cost_{i}
    \label{eq:asset_cost}
\end{align} 
For example the total OPEX, $\opexgeneration = \sum_{s} \opexgeneration_{s}$, is the combined OPEX of all generators $s$.

In a cost-optimal setup with minimized $\totalcost$, the Locational Marginal Price (LMP) describes the price for an incremental increase of electricity demand $\demand$ at node $n$. It is given by the derivative of the total system cost $\totalcost$ with respect to the local demand $\demand$
\begin{align}
\lmp = \pdv{\,\totalcost}{\,\demand}
\label{eq:lmp}
\end{align}
% 
\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{price_decomposition.png}
\caption{Schematic decomposition of the Locational Market Price $\lmp$. In power system model with optimal long-term operation and planning, the total system costs $\totalcost$ split into different cost terms, \ie OPEX and CAPEX for production and transmission and possibly other expenditures. }
\label{fig:price_decomposition}
\end{figure}
% 
This leads to a nodal pricing where over the span of optimized timesteps $t$, the system costs are partially or totally payed back by the consumers 
\begin{align}
\totalcost - \remainingcost &=  \sum_{n,t} \lmp \, \demand
\label{eq:total_cost_reduced}
\end{align}
depending on the costs $\remainingcost$ which are independent of the nodal demand  
\begin{align}
 \pdv{\remainingcost}{\demand} = 0
\end{align}
% 
Generally speaking, the cost term $\remainingcost$, not covered by the consumers, results from additional demands on the network design, such as capacity expansion limits or minimum share of one technology in the power mix. However, in most cases, where $\remainingcost \ll  \totalcost$, these play a minor role. From feeding \cref{eq:total_cost} into \cref{eq:lmp} it follows naturally that the LMP splits into different cost terms assigned to OPEX and CAPEX for production and transmission (and other possible expenditures). This relation, which we schematically show in \cref{fig:price_decomposition}, was already shown in extensive investigations of the LMP \cite{schweppe_spot_1988}.

However the question of how the LMP can be decomposed into contributions of single cost terms $\cost_{i}$ associated with asset $i$ remains unanswered. This work aims at presenting and illustrating a an intuitive, peer-to-peer cost allocation including all network assets.

@@

Let $\cost_{i}$ denote a general cost term associated with asset $i$. Consider a long-term equilibrium in a power system with perfect competition, then, according to the zero-profit condition, each cost term $\cost_{i}$ can be considered as a cost-weighted sum of the operational state $s_{i,t}$ of asset $i$, \ie
\begin{align}
    \cost_{i} = \sum_t  \costfactor \, \state
    \label{eq:cost_decomposition}
\end{align}
where $c_{i,t}$ denotes a general cost factor in \euro/MW. 
For OPEX the cost factor $\costfactor$ is simply given by the marginal operational price $o_i$ for asset $i$. However for CAPEX, $c_{i,t}$ is a composition of shadow prices $\mu_i$, given by the KKT variables of the capacity limit constraints. As we will show later the composition of $\mu_i$ must be determined for each cost term individually.

% where $c_{i,t}$ is either a fixed operational price or a composition of shadowprices
Following the implications of \cref{eq:lmp,eq:total_cost_reduced},  we define the cost $\allocatecost$ that consumers at bus $n$ have pay to asset $i$ at time $t$ as  
\begin{align}
    \allocatecost = \costfactor \,\pdv{\state}{\demand} \demand
\end{align}
Therefore, $\allocatecost$ indicates the contribution of $\demand$ to the cost term $\cost_{i}$. It heavily relies on the derivative of the operational state with respect to the nodal demand $\partial \state / \partial \demand$. Without further specifying, we define the derivative multiplied with the nodal demand as 
\begin{align}
    \allocatestate = \pdv{\state}{\demand} \demand
    \label{eq:allocate_state}
\end{align}    
This quantity may be interpreted as the amount of power that asset $i$ supplies demand $\demand$ with.
From the natural fact that the sum of all contributions must return the cost term, 
\begin{align}
    \cost_i = \sum_{n,t} \allocatecost
\end{align}
it follows that $\allocatestate$ must fulfill
\begin{align}
    \state = \sum_n \allocatestate
    \label{eq:allocate_state_constraint}
\end{align}
Finally, the total contribution from node $n$ at time $t$ to the cost term $\cost$ amounts 
\begin{align}
    \cost_{n,t} = \sum_i \allocatecost
\end{align}
\end{subequations}

\begin{table*}[t]
    \begin{center}
        \begin{tabular}{l|c|c|c|c|c}
        & $i$ & $\cost$ & $\cost_i$  & $\costfactor$ & $\state$  \\
        \toprule 
        OPEX Production & $s$ & $\opexgeneration$ & $\sum_{t} \operationalpricegeneration \, \generation$   & $\operationalpricegeneration$ & $\generation$ \\  
        OPEX Transmission  & $\ell$ & $\opexflow$ & $\sum_{t} \operationalpriceflow \, |\flow|  $ & $\operationalpriceflow$ & $|\flow|$ \\  
        OPEX Storage & $r$  & $\opexstorage$ & $\sum_{t} \operationalpricestorage \, \storagedispatch$ &  $\operationalpricestorage$ & $\storage$ \\
        \midrule   
        CAPEX Production & $s$ & $\capexgeneration$ & $ \capitalpricegeneration \capacitygeneration$ & $\muuppergeneration$ & $\generation$ \\
        CAPEX Transmission & $\ell$ & $\capexflow$ & $ \capitalpriceflow \capacityflow$ & $\left(\muupperflow - \mulowerflow \right)$ & $\flow$ \\
        CAPEX Storage & $r$ & $\capexstorage$ & $ \capitalpricestorage \capacitystorage$ & $ (\efficiencydispatch )^{-1} \mustateofcharge - \mulowerstoragedispatch$ & $\storage$ \\
        \midrule
        Emission Cost & $s$ & $\emissioncost$ & $ \emissionprice \, \emission \, \generation$ & $\emissionprice \,\emission$ & $\generation$ \\
        % \bottomrule   
    \end{tabular}
    \end{center}
    \caption{Mapping of different cost terms to the cost allocation scheme given in \cref{eq:general_scheme}. These include OPEX \& CAPEX for production, transmission and storage assets in the network, as well as a cost term for the total Green House Gas (GHG) emissions.}
    \label{tab:cost_allocation_map}
\end{table*}

Now, assume a network with generators $s$, transmission lines $\ell$ and storage units $r$. Each asset $i = \{s, \ell, r\}$ adds an term for OPEX and a term for CAPEX to the total system cost $\totalcost$. The operational price for an asset $i$ is given by $o_i$, the capital price for one unit capacity expansion by $c_i$. For example the OPEX occasioned by generator $s$ is given by 
\begin{align}
    \opexgeneration_s = \sum_t \operationalpricegeneration \, \generation 
    \label{eq:opexgeneration}
\end{align}
where $\generation$ denotes its production at time $t$. It becomes clear that \cref{eq:opexgeneration} matches the form of \cref{eq:cost_decomposition} which allows us to use the above presented scheme in \cref{eq:general_scheme} for a full cost allocation of OPEX on production. As a result we obtain the contribution of $\demand$ to the OPEX at generator $s$, given by 
\begin{align}
    \allocateopex[n \rightarrow s] &= 
   \operationalpricegeneration \,  \allocategeneration
\label{eq:allocate_opexGeneration_detailed}
\end{align}
The quantity  
\begin{align}
 \allocategeneration = \pdv{\generation}{\demand} \, \demand
 \label{eq:allocate_peer}
\end{align}
can be considered as a power allocation which indicates the power that is produced by generator $s$ and consumed at node $n$ at time $t$.
It follows naturally that, in the same manner, the scheme can be applied to the OPEX on the flow $\flow$ and storage unit dispatch $\storagedispatch$. As we assume a bidirectional flow on line $\ell$, the OPEX is set proportional to the absolute value of the flow. The upper section in \cref{tab:cost_allocation_map} shows the mapping of variables to \cref{eq:general_scheme} in order to define the full OPEX allocation. Note, in all cases, the allocation $\allocatestate$ remains undefined which we will resolve later.

For the CAPEX allocation, it becomes crucial to look at the individual relations between operational state $\state$ and the capacity limit. All quantities for the CAPEX allocation, which we discuss in detail in the following, are summarized in the middle section of \cref{tab:cost_allocation_map}.

@@

The nominal capacity $\capacitygeneration$ constrains the generation $\generation$ in the form of 
\begin{align}
\generation - \generationpotential \capacitygeneration  &\le 0 \resultsin{\muuppergeneration} \Forall{s,t} 
\label[constraint]{eq:upper_generation_capacity_constraint}\\ 
- \generation &\le 0 \resultsin{\mulowergeneration} \Forall{s,t} 
\label[constraint]{eq:lower_generation_capacity_constraint}
\end{align}
where $\generationpotential \in \left[ 0,1\right]$ is the capacity factor for renewable generators. At a cost-optimum, these two constraints yield the shadow prices $\muuppergeneration$ and $\mulowergeneration$.  As shown in \cite{brown_decreasing_2020} and in detailed in \cref{sec:zero_profit_generation}, over the whole time span, the CAPEX for generator $s$ is payed back by the production $\generation$ times the shadow price $\muuppergeneration$, 
\begin{align}
 \capexgeneration_s = \capitalpricegeneration \capacitygeneration = \sum_t \muuppergeneration \,  \generation 
 \label{eq:no_profit_capex_generation}
\end{align}
This representation connects the CAPEX with the operational state of generator $s$ and matches the needed form in \cref{eq:cost_decomposition} for using the allocation scheme. The resulting the CAPEX allocation is given by
\begin{align}
 \allocatecapexgeneration = \muuppergeneration \, \allocategeneration
 \label{eq:allocate_capexGeneration_detailed}
\end{align}
How does this allocation behave? According to the polluter pays principle, it differentiates between consumers who are `responsible` for investments and those who are not. If $\muuppergeneration$ (in literature often denoted as the Quality of Supply) is bigger than zero, the upper Capacity \cref{eq:upper_generation_capacity_constraint} is binding. Thus it is these times steps which push investments in $\capacitygeneration$. If $\muuppergeneration = 0$, the generation $\generation$ is not bound and investments are not necessary. 
When summing over all CAPEX payments to generator $s$ in \cref{eq:allocate_capexGeneration_detailed} and using \cref{eq:allocate_state_constraint,eq:no_profit_capex_generation}, we see that each generator retrieves exactly the cost that were spent to build the capacity $\capacitygeneration$.

@@

The transmission capacity $\capacityflow$ limits the flow $\flow$ in both directions,
\begin{align}
\flow - \capacityflow &\le 0 \resultsin{\muupperflow} \Forall{\ell,t} 
\label[constraint]{eq:upper_flow_capacity_constraint} \\
- \flow - \capacityflow &\le 0 \resultsin{\mulowerflow} \Forall{\ell,t} 
\label[constraint]{eq:lower_flow_capacity_constraint}
\end{align}
which yield the shadow prices $\muupperflow$ and $\mulowerflow$. Again, we use the result of \cite{brown_decreasing_2020} (for details see \cref{sec:zero_profit_flow}) which derives that over the whole time span, the investment in line $\ell$ is payed back by the shadow prices times the flow 
\begin{align}
\capexflow_\ell = \capitalpriceflow \capacityflow = \sum_{t} \left( \muupperflow - \mulowerflow \right)  \flow 
\label{eq:no_profit_capex_flow}
\end{align}
Again, we follow the scheme in \cref{eq:general_scheme} which  defines the CAPEX allocation as 
\begin{align}
    \allocatecapexflow &=  
   \left( \muupperflow - \mulowerflow\right) \, \allocateflow
   \label{eq:allocate_capexFlow_detailed}
\end{align}
The quantity 
\begin{align}
 \allocateflow =  \pdv{\flow}{\demand}\, \demand
\end{align}
can be interpreted as the flow that the demand at $\left(n,t\right)$ causes on line $\ell$.
% According to \cref{eq:allocate_state_constraint} it has to fulfill 
% \begin{align}
%  \flow = \sum_{n} \allocateflow
% \label{eq:allocate_flow_constraint}
% \end{align}
The  shadowprices $\muupperflow$ and $\mulowerflow$ again can be seen as a measure for necessity of transmission investments at $\ell$ at time $t$. Hence, the definition of $\allocatecapexflow$ states that consumers, which retrieve power flowing on congested lines, yielding a bound \cref{eq:upper_flow_capacity_constraint} or \eqref{eq:lower_flow_capacity_constraint}, pay compensations for the resulting investments at $\ell$. Again the sum of all CAPEX payments to line $\ell$ equals the total CAPEX spent. This is seen when summing \cref{eq:allocate_capexFlow_detailed} over all buses and time steps and using \cref{eq:allocate_state_constraint,eq:no_profit_capex_flow}

@@

The presented cost allocation suits for any type of topology and network setup. But so far, the question of how $\allocatestate$ for generators $s$, lines $\ell$ and storages $r$ are defined was left open. We recap that all rely on derivatives of $\generation$, $\flow$ and $\storagedispatch$  with respect to the nodal demand $\demand$ (see \cref{eq:allocate_state_constraint}). \\

@@

\subsection{LMP from Optimization}
The nodal balance constraint ensures that the amount of power that flows into a bus equals the power that flows out of a bus, thus reflects the Kirchhoff Current Law (KCL)
\begin{align}
    \sum_l \incidence \, \flow  - \nodalgeneration + \nodaldemand &= 0 \resultsin{\lmp} \Forall{n,t}
    \label[constraint]{eq:nodal_balance_lin}
\end{align}
Its shadow price mirrors the Locational Marginal Prizes (LMP) $\lmp$ per bus and time step. In a power market this is the \euro/\megawatthour-price which a consumer has to pay.\\

>>>>

% general symbols
\newcommand{\state}{s_{i,t}}
\newcommand{\capacity}{S_{i}}
\newcommand{\costfactor}{\gamma_{i,t}}

@@

% storage
\newcommand{\storage}{g_{r,t}}
\newcommand{\storagedispatch}{\storage^\text{dis}}
\newcommand{\storagecharge}{\storage^\text{sto}}
\newcommand{\storagesoc}{\storage^\text{ene}}
\newcommand{\storageprevioussoc}{g_{r,t-1}^\text{ene}}

@@

\newcommand{\operationalpricestorage}{o_r}
\newcommand{\capitalpricestorage}{c_r}
\newcommand{\capacitystorage}{G_r}
\newcommand{\capacitystorageupper}{\bar{G}_r}
\newcommand{\capacitystoragelower}{\ubar{G}_r}
\newcommand{\mulowerstorage}{\ubar{\mu}_{r,t}}
\newcommand{\muupperstorage}{\bar{\mu}_{r,t}}
\newcommand{\mulowerstoragedispatch}{\ubar{\mu}_{r,t}^\text{dis}}
\newcommand{\muupperstoragedispatch}{\bar{\mu}_{r,t}^\text{dis}}
\newcommand{\mulowerstoragecharge}{\ubar{\mu}_{r,t}^\text{sto}}
\newcommand{\muupperstoragecharge}{\bar{\mu}_{r,t}^\text{sto}}
\newcommand{\mulowerstoragesoc}{\ubar{\mu}_{r,t}^\text{ene}}
\newcommand{\muupperstoragesoc}{\bar{\mu}_{r,t}^\text{ene}}
\newcommand{\muupperstoragenom}{\bar{\mu}^\text{nom}_r}
\newcommand{\mulowerstoragenom}{\ubar{\mu}^\text{nom}_r}
\newcommand{\mustateofcharge}{\lambda^\text{ene}_{r,t}}
\newcommand{\munextstateofcharge}{\lambda^\text{ene}_{r,t+1}}

@@

% allocation quantities
\newcommand{\allocategeneration}[1][s, n]{A_{#1,t}}
\newcommand{\allocatestoragedispatch}[1][r, n]{A_{#1,t}}
\newcommand{\allocatepeer}[1][m \rightarrow n]{A_{#1,t}}
\newcommand{\allocateflow}[1][n]{A_{\ell,#1,t}}
\newcommand{\allocatetransaction}[1][s \rightarrow n]{A_{#1,\ell,t}}
\newcommand{\allocatestate}[1][i, n]{A_{#1,t}}
\newcommand{\allocatecost}[1][n \rightarrow i]{\mathcal{C}_{#1, t}}
\newcommand{\allocatecapexgeneration}[1][n \rightarrow s]{\mathcal{C}^{G}_{#1,t}}
\newcommand{\allocatecapexflow}[1][n \rightarrow \ell]{\mathcal{C}^{F}_{#1,t}}
\newcommand{\allocatecapexstorage}[1][n \rightarrow r]{\mathcal{C}^{E}_{#1,t}}
\newcommand{\allocateopex}[1][n \rightarrow s]{\mathcal{O}_{#1,t}}
\newcommand{\allocateemissioncost}[1][n \rightarrow s]{\mathcal{E}_{#1,t}}

@@

In long-term operation and investment planning models, the total costs $\totalcost$ of a power system is the sum of mutliple cost terms. Typically, these include operational expenditures (OPEX) $\opexgeneration$ for generators, expenditures for emissions $\emissioncost$, capital expenditures (CAPEX) for generators $\capexgeneration$, CAPEX for the transmission system $\capexflow$ and possible other terms, \ie
\begin{align}
\totalcost &= \opexgeneration + \emissioncost +  \capexgeneration +  \capexflow + ...
\label{eq:total_cost}
\end{align}
In turn, each of these terms $\mathcal{C} = \{\opexgeneration, \emissioncost, \capexgeneration, \capexflow, ...\}$ consists of cost associated to the asset $i$ in the system, 
\begin{align}
    \cost = \sum_{i} \cost_{i}
    \label{eq:asset_cost}
\end{align} 
where an asset describes any component of the network, such as a generator, line, energy storage etc. %For example the total OPEX, $\opexgeneration = \sum_{s} \opexgeneration_{s}$, is the combined OPEX of all generators $s$.

In a cost-optimal setup with minimized $\totalcost$, the Locational Marginal Price (LMP) describes the price for an incremental increase of electricity demand $\demand$ at node $n$. It is given by the derivative of the total system cost $\totalcost$ with respect to the local demand $\demand$
\begin{align}
\lmp = \pdv{\,\totalcost}{\,\demand}
\label{eq:lmp}
\end{align}
% 
\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{price_decomposition.png}
\caption{Schematic decomposition of the Locational Market Price $\lmp$. In power system model with optimal long-term operation and planning, the total system costs $\totalcost$ split into different cost terms, \ie OPEX and CAPEX for production and transmission and possibly other expenditures. }
\label{fig:price_decomposition}
\end{figure}
% 
This leads to a nodal pricing where over the span of optimized timesteps $t$, the system costs are partially or totally payed back by the consumers 
\begin{align}
\totalcost - \remainingcost &=  \sum_{n,t} \lmp \, \demand
\label{eq:total_cost_reduced}
\end{align}
depending on the costs $\remainingcost$ which are independent of the nodal demand  
\begin{align}
 \pdv{\remainingcost}{\demand} = 0
\end{align}
% 
Generally speaking, the cost term $\remainingcost$, not covered by the consumers, results from additional demands on the network design, such as capacity expansion limits or minimum share of one technology in the power mix. However, in most cases, where $\remainingcost \ll  \totalcost$, these play a minor role.

From feeding \cref{eq:total_cost} into \cref{eq:lmp} it follows naturally that the LMP splits into contribution to the above mentioned cost terms. This relation, which we schematically show in \cref{fig:price_decomposition}, was already shown in extensive investigations of the LMP \cite{schweppe_spot_1988}. However the question of how the LMP can be decomposed into contributions of single cost terms $\cost_{i}$ associated with asset $i$ remains unanswered. This work aims at presenting and illustrating a an intuitive, peer-to-peer cost allocation including all network assets.

@@

Let $\cost_{i}$ denote a general cost term associated with asset $i$. Consider a long-term equilibrium in a power system with perfect competition, then, according to the zero-profit condition, each cost term $\cost_{i}$ can be considered as a cost-weighted sum of the operational state $s_{i,t}$ of asset $i$, \ie
\begin{align}
    \cost_{i} = \sum_t  \costfactor \, \state
    \label{eq:cost_decomposition}
\end{align}
where $\costfactor$ denotes a cost factor in \euro/MW. 
If $\cost_i$ describes the OPEX occasioned by asset $i$, the cost factor $\costfactor$ is simply given by the marginal operational price $o_i$. However, as we will show later, if it describes the CAPEX of asset $i$, $\costfactor$ is a composition of shadow prices $\mu_{i,t}$ given at the optimum. %As we will show later the composition of $\mu_{i,t}$ must be determined for each type of asset individually.

% where $\costfactor$ is either a fixed operational price or a composition of shadowprices
Following the implications of \cref{eq:lmp,eq:total_cost_reduced},  we define the cost $\allocatecost$ that consumers at bus $n$ have pay to asset $i$ at time $t$, in order to compensate for $\cost_i$, as  
\begin{align}
    \allocatecost = \costfactor \,\pdv{\state}{\demand} \demand
    \label{eq:cost_allocation}
\end{align}
The product on the right hand side 
\begin{align}
    \allocatestate = \pdv{\state}{\demand} \demand
    \label{eq:state_allocation}
\end{align}    
may be interpreted as the amount of power that asset $i$ supplies demand $\demand$ with. It heavily relies on the derivative of the operational state with respect to the nodal demand $\partial \state / \partial \demand$.
From the natural fact that the sum of all contributions must return the cost term, 
\begin{align}
    \cost_i = \sum_{n,t} \allocatecost
    \label{eq:cost_payback}
\end{align}
it follows that $\allocatestate$ must fulfill
\begin{align}
    \state = \sum_n \allocatestate
    \label{eq:state_allocation_constraint}
\end{align}
The last equation states that all power produced or processed by asset $i$ must be totally consumed by the network demand $\demand$.  
Finally, the total contribution from node $n$ at time $t$ to the cost term $\cost$ amounts 
\begin{align}
    \cost_{n,t} = \sum_i \allocatecost
\end{align}
\end{subequations}

\begin{table*}[t]
    \begin{center}
        \begin{tabular}{l|c|c|c|c|c}
        & $i$ & $\cost$ & $\cost_i$  & $\costfactor$ & $\state$  \\
        \toprule 
        OPEX Production & $s$ & $\opexgeneration$ & $\sum_{t} \operationalpricegeneration \, \generation$   & $\operationalpricegeneration$ & $\generation$ \\  
        OPEX Transmission  & $\ell$ & $\opexflow$ & $\sum_{t} \operationalpriceflow \, |\flow|  $ & $\operationalpriceflow$ & $|\flow|$ \\  
        OPEX Storage & $r$  & $\opexstorage$ & $\sum_{t} \operationalpricestorage \, \storagedispatch$ &  $\operationalpricestorage$ & $\storage$ \\
        \midrule   
        CAPEX Production & $s$ & $\capexgeneration$ & $ \capitalpricegeneration \capacitygeneration$ & $\muuppergeneration$ & $\generation$ \\
        CAPEX Transmission & $\ell$ & $\capexflow$ & $ \capitalpriceflow \capacityflow$ & $\left(\muupperflow - \mulowerflow \right)$ & $\flow$ \\
        CAPEX Storage & $r$ & $\capexstorage$ & $ \capitalpricestorage \capacitystorage$ & $ \muupperstoragedispatch - \mulowerstoragedispatch  + (\efficiencydispatch )^{-1} \mustateofcharge $ & $\storage$ \\
        \midrule
        Emission Cost & $s$ & $\emissioncost$ & $ \emissionprice \, \emission \, \generation$ & $\emissionprice \,\emission$ & $\generation$ \\
        % \bottomrule   
    \end{tabular}
    \end{center}
    \caption{Mapping of different cost terms to the cost allocation scheme given in \cref{eq:general_scheme}. These include OPEX \& CAPEX for production, transmission and storage assets in the network, as well as a cost term for the total Green House Gas (GHG) emissions.}
    \label{tab:cost_allocation_map}
\end{table*}

Now, assume a network with generators $s$, transmission lines $\ell$ and storage units $r$. Each asset $i = \{s, \ell, r\}$ adds an term for OPEX and a term for CAPEX to the total system cost $\totalcost$. The operational price for an asset $i$ is given by $o_i$, the capital price for one unit capacity expansion by $c_i$. For example the OPEX occasioned by generator $s$ is given by 
\begin{align}
    \opexgeneration_s = \sum_t \operationalpricegeneration \, \generation 
    \label{eq:opexgeneration}
\end{align}
where $\generation$ denotes its power generation at time $t$. As \cref{eq:opexgeneration} matches the form of \cref{eq:cost_decomposition} which allows us to use the above presented scheme in \cref{eq:general_scheme}. As a result we obtain 
\begin{align}
    \allocateopex[n \rightarrow s] &= 
   \operationalpricegeneration \,  \allocategeneration
\label{eq:allocate_opexGeneration_detailed}
\end{align}
which is the contribution of $\demand$ to the OPEX at generator $s$.
The quantity  
\begin{align}
 \allocategeneration = \pdv{\generation}{\demand} \, \demand
 \label{eq:allocate_peer}
\end{align}
can be considered as the power that is produced by generator $s$ and consumed at node $n$ at time $t$. \\
In the same manner, we can follow the scheme allocate OPEX for flow $\flow$ and storage unit dispatch $\storagedispatch$. As we assume a bidirectional flow on line $\ell$, the OPEX is set proportional to the absolute value of the flow. The upper section in \cref{tab:cost_allocation_map} shows the mapping of variables to \cref{eq:general_scheme} in order to define the full OPEX allocation.

For the CAPEX allocation, it becomes crucial to look at the individual relations between operational state $\state$ and the capacity limit. All quantities for the CAPEX allocation, which we now discuss in detail in the following, are summarized in the middle section of \cref{tab:cost_allocation_map}.

@@

The nominal capacity $\capacitygeneration$ constrains the generation $\generation$ in the form of 
\begin{align}
\generation - \generationpotential \capacitygeneration  &\le 0 \resultsin{\muuppergeneration} \Forall{s,t} 
\label[constraint]{eq:upper_generation_capacity_constraint}\\ 
- \generation &\le 0 \resultsin{\mulowergeneration} \Forall{s,t} 
\label[constraint]{eq:lower_generation_capacity_constraint}
\end{align}
where $\generationpotential \in \left[ 0,1\right]$ is the capacity factor for renewable generators. At a cost-optimum, these two constraints yield the shadow prices $\muuppergeneration$ and $\mulowergeneration$.  As shown in \cite{brown_decreasing_2020} and in detailed in \cref{sec:zero_profit_generation}, over the whole time span, the CAPEX for generator $s$ is payed back by the production $\generation$ times the shadow price $\muuppergeneration$, 
\begin{align}
 \capexgeneration_s = \capitalpricegeneration \capacitygeneration = \sum_t \muuppergeneration \,  \generation 
 \label{eq:no_profit_capex_generation}
\end{align}
This representation connects the CAPEX with the operational state of generator $s$, \ie matches the form in \cref{eq:cost_decomposition} allows for using the cost allocation scheme. The resulting the CAPEX allocation is given by
\begin{align}
 \allocatecapexgeneration = \muuppergeneration \, \allocategeneration
 \label{eq:allocate_capexGeneration_detailed}
\end{align}
How does this allocation behave? According to the polluter pays principle, it differentiates between consumers who are `responsible` for investments and those who are not. If $\muuppergeneration$ (in literature often denoted as the Quality of Supply) is bigger than zero, the upper Capacity \cref{eq:upper_generation_capacity_constraint} is binding. Thus it is these times steps which push investments in $\capacitygeneration$. If $\muuppergeneration = 0$, the generation $\generation$ is not bound and investments are not necessary. 
When summing over all CAPEX payments to generator $s$ in \cref{eq:allocate_capexGeneration_detailed} , we can use \cref{eq:state_allocation_constraint,eq:no_profit_capex_generation} to see that each generator retrieves exactly the cost that were spent to build the capacity $\capacitygeneration$.

@@

The transmission capacity $\capacityflow$ limits the flow $\flow$ in both directions,
\begin{align}
\flow - \capacityflow &\le 0 \resultsin{\muupperflow} \Forall{\ell,t} 
\label[constraint]{eq:upper_flow_capacity_constraint} \\
- \flow - \capacityflow &\le 0 \resultsin{\mulowerflow} \Forall{\ell,t} 
\label[constraint]{eq:lower_flow_capacity_constraint}
\end{align}
which yield the shadow prices $\muupperflow$ and $\mulowerflow$. Again, we use the result of \cite{brown_decreasing_2020} (for details see \cref{sec:zero_profit_flow}) which derives that over the whole time span, the investment in line $\ell$ is payed back by the shadow prices times the flow 
\begin{align}
\capexflow_\ell = \capitalpriceflow \capacityflow = \sum_{t} \left( \muupperflow - \mulowerflow \right)  \flow 
\label{eq:no_profit_capex_flow}
\end{align}
From here, we follow the scheme in \cref{eq:general_scheme} which finally defines the CAPEX allocation as 
\begin{align}
    \allocatecapexflow &=  
   \left( \muupperflow - \mulowerflow\right) \, \allocateflow
   \label{eq:allocate_capexFlow_detailed}
\end{align}
The quantity 
\begin{align}
 \allocateflow =  \pdv{\flow}{\demand}\, \demand
\end{align}
can be interpreted as the flow that the demand at node $n$ and time $t$ causes on line $\ell$.
% According to \cref{eq:state_allocation_constraint} it has to fulfill 
% \begin{align}
%  \flow = \sum_{n} \allocateflow
% \label{eq:allocate_flow_constraint}
% \end{align}
The  shadowprices $\muupperflow$ and $\mulowerflow$ again can be seen as a measure for necessity of transmission investments at $\ell$ at time $t$. Hence, the definition of $\allocatecapexflow$ states that consumers, which retrieve power flowing on congested lines, yielding a bound \cref{eq:upper_flow_capacity_constraint} or \eqref{eq:lower_flow_capacity_constraint}, pay compensations for the resulting investments at $\ell$. Again the sum of all CAPEX payments to line $\ell$ equals the total CAPEX spent. This is seen when summing \cref{eq:allocate_capexFlow_detailed} over all buses and time steps and using \cref{eq:state_allocation_constraint,eq:no_profit_capex_flow}

@@

The presented cost allocation suits for any type of topology and network setup. But so far, the question of how $\allocatestate$ for generators $s$, lines $\ell$ and storages $r$ are defined was left open. We recap that all rely on derivatives of $\generation$, $\flow$ and $\storagedispatch$  with respect to the nodal demand $\demand$ (see \cref{eq:state_allocation_constraint}). \\

@@

\subsection{LMP from Optimization}
The nodal balance constraint ensures that the amount of power that flows into a bus equals the power that flows out of a bus, thus reflects the Kirchhoff Current Law (KCL). Alternativaly, we can the demand $\demand$ has to be supplied by the attached assets,  
\begin{align}
    \sum_i \incidenceasset \state  &=  \demand 
    % \sum_l \incidence \, \flow  - \nodalgeneration + \nodaldemand &= 0
     \resultsin{\lmp} \Forall{n,t}
    \label[constraint]{eq:nodal_balance_lin}
\end{align}
where $\incidenceasset$ is +1 if $i$ is attached to $n$ and a positive operation $\state$ delivers power to $n$, -1 if is attached to $n$ and a positive operation retrieves power from $n$ and zero else (note that for lines this results in the negative of the conventional Incidence Matrix).  
The shadow price of the nodal balance constraint mirrors the Locational Marginal Prizes (LMP) $\lmp$ per bus and time step. In a power market this is the \euro/\megawatthour-price which a consumer has to pay.\\

\subsection{Full Lagrangian}
\label{sec:full_lagrangian}
The Lagrangian for the investment model can be condensed to the following expression
